---
title: "Mapping between different Genome references"
output:
  BiocStyle::html_document:
    toc: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Ensembl and UCSC mapping}
%\VignettePackage{Pbase}
-->

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package:** [`Pbase`](http://bioconductor.org/packages/devel/bioc/html/Pbase.html)<br />
**Author:** [Laurent Gatto](http://cpu.sysbiol.cam.ac.uk/)<br />
**Compilation date:** `r Sys.Date()`<br />

This vignette described how to convert coordinates between different
genome references. We will use transcript `ENST00000373316` and GRCh38
and GRCh37 as working example.

```{r bm} 
tr <- "ENST00000373316"
``` 

## GRCh38

Here is use `r Biocpkg("Gviz")` to query the latest Ensembl biomart
and extract the transcript of interest.

```{r h38}
suppressMessages(library("Gviz"))
suppressMessages(library("biomaRt"))

h38 <- useMart("ensembl", "hsapiens_gene_ensembl")
tr38 <- BiomartGeneRegionTrack(biomart = h38,
                               transcript = tr)
tr38 <- split(tr38, transcript(tr38))
tr38 <- ranges(tr38[[tr]])
tr38
```

Note the starting position of the transcript is `r min(start(tr38))`.

## GRCh37

Below, I repeat the same operation without using my own ens Mart
instance. As far as I understand, Gviz queries the UCSC genome
reference by default. 

```{r h37}
h37 <- useMart(host = "feb2014.archive.ensembl.org",
                biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "hsapiens_gene_ensembl")
tr37 <- BiomartGeneRegionTrack(biomart = h37,
                               transcript = tr)
tr37 <- split(tr37, transcript(tr37))
tr37 <- ranges(tr37[[tr]])
tr37
```

Note the starting position of the transcript is `r min(start(tr37))`.

These differences seem to stem from different genome builds. **Ensembl
release 78** uses **GRCh38**, while **UCSC** uses **GRCh37**. Indeed,
`r Biocpkg("Gviz")` sets the Ensembl biomart server to `Feb.2014`
`GRCh37.p13`.

## Coordinates conversion

We will use the coordinate mapping infrastructure described in the
[January 2015 Bioconductor Newletter](http://www.bioconductor.org/help/newsletters/2015_January/#coordinate-mapping)
and the
[Changing genomic coordinate systems with rtracklayer::liftOver](http://bioconductor.org/help/workflows/liftOver/)
workflow.

First, we query `r Biocpkg("AnnotationHub")` for a chain file to
perform the operation we want.

```{r chain}
library("AnnotationHub")
hub <- AnnotationHub()
query(hub, 'hg19ToHg38')
chain <- query(hub, 'hg19ToHg38')[[1]]
```

The `liftOver` function from the `r Biocpkg("rtracklayer")` package
will use the chain and translate the coordinates of a `GRanges` object
into a new `GRangesList` object.

```{r}
library("rtracklayer")
res <- liftOver(tr37, chain)
res <- unlist(res)

## set annotation
names(res) <- NULL
genome(res) <- "hsapiens_gene_ensembl"

all.equal(res, tr38)
```

## Session information

```{r si}
sessionInfo()
```
