---
title: "Pbase example data"
author: "Laurent Gatto - <lg390@cam.ac.uk>"
output:
  html_document:
    toc: true
    theme: united
---

```{r, env, echo=FALSE, message=FALSE, warning=FALSE}
library("Pbase")
```

The original data is a 10 fmol
[Peptide Retention Time Calibration Mixture](http://www.piercenet.com/product/peptide-retention-time-calibration-mixture)
spiked into 50 ng HeLa background acquired on a Thermo Orbitrap Q
Exactive instrument. A restricted set of high scoring human proteins
from the UniProt release 2014_06 were searched using the MSGF+ engine.

## The fasta database

```{r, fa, cache=TRUE}
library("Biostrings")
fafile <- system.file("extdata/HUMAN_2014_06-09prots.fasta", package = "Pbase")
fa <- readAAStringSet(fafile)
fa
```

## The PSM data

```{r, psm, cache=TRUE}
library("mzID")
idfile <- system.file("extdata/Thermo_Hela_PRTC_1_selected.mzid", package = "Pbase")
id <- flatten(mzID(idfile))
dim(id)
head(id)
```

## The Proteins object

```{r, p, cache=TRUE}
library("Pbase")
p <- Proteins(fafile)
p <- addIdentificationData(p, idfile)
p
```

A `Proteins` object is composed of a set of protein sequences
accessible with the `aa` accessor as well as an optional set of
peptides features that are mapped as coordinates along the proteins,
available with `pranges`. The actual peptide sequences can be extraced
with `pfeatures`.

```{r, paccess}
aa(p)
pranges(p)
pfeatures(p)
```

A Proteins instance is further described by general
`metadata`. Protein sequence and peptide features annotations can be
accessed with `ametadata` and `pmetadata` (or `acols` and `pcols`)
respectively.

```{r, metadata}
metadata(p)
head(acols(p))
head(pcols(p))
```

Specific proteins can be extracted by index of name using
`[` and proteins and their peptide features can be plotted
with the default plot method.

```{r, pplot, fig.align='center', cache=TRUE}
seqnames(p)
plot(p[c(1,9)])
```

More details can be found in `?Proteins`.

## Session information

```{r, si}
sessionInfo()
```
